\documentclass[a4paper]{article}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{times}
\usepackage{amssymb}
\usepackage{mathtools}	% pulls in amsmath
	\mathtoolsset{centercolon}
\usepackage{tikz}
	\usetikzlibrary{automata}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{amsxtra}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{semantic}
	\reservestyle{\declarevars}{\texttt}
	\reservestyle{\declareops}{\texttt}
	\reservestyle{\declarestates}{\text}
\usepackage{color}
\usepackage{listings}
\usepackage{mathtools}

\newtheorem{theorem}{Theorem}

\newtheorem{myexample}{\textbf{Example}}
\newtheorem{mylemma}{\textbf{Lemma}}
\newtheorem{myproof}{\textbf{Proof}}
\newtheorem{myinvariant}{\textbf{Invariant}}
\newtheorem{mytheorem}{\textbf{Theorem}}
\newtheorem{mycorollary}{\textbf{Corollary}}
\newtheorem{myapproach}{Approach}
\newtheorem{myproperty}{Property}
\newtheorem{mydefinition}{Definition}

\newtheorem{mycase}{Case}

\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\small,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
 % frame=single,                    % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  columns=fullflexible,	% not monospace
  keywordstyle=\color{blue},       % keyword style
  language=Octave,                 % the language of the code
  morekeywords={then, end, and, or, assign, increment, decrement, jump, jump_if, store, *, +},            % if you want to add more keywords to the set
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=5pt,                   % how far the line-numbers are from the code
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{mymauve},     % string literal style
  tabsize=2,                       % sets default tabsize to 2 spaces
  title=\lstname,                  % show the filename of files included with \lstinputlisting; also try caption instead of title
  mathescape
}

\newcommand*{\floor}[1]{\left\lfloor{#1}\right\rfloor}
\newcommand{\any}{{\rule[-.2ex]{1ex}{.4pt}}}	% Less hideous than \_.
\newcommand{\RR}{\mathbb{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RP}{\RR_{\ge 0}}
\newcommand*{\dave}[1]{{\color{red}\textbf{PDS: #1}}}
\newcommand{\ie}{\emph{i.e.,} }
\newcommand{\eg}{\emph{e.g.,} }
\usepackage{hyperref}
\newcommand*{\figref}[1]{\hyperref[#1]{Figure~\ref*{#1}}}

\title{Exercise Sheet 3---Algorithms and Data Structures}
\author{Felipe Cerqueira \\ 2547787 \and David Swasey \\ 2542105}

\begin{document}

\maketitle

Tutorial time: Monday 14:00

\section{Exercise 1 (10 pts)}

\subsection{Part a}
Consider the following pivot-selection strategies for the quicksort algorithm for a sequence $a_0, \ldots, a_{n-1}$ of $n$ distinct integers:
\begin{itemize}
	\item Pick $a_m$ with $m := \floor{n/2}$ as pivot.
	\item Look at positions $a_0$, $a_{n-1}$, and $a_m$ (as above) and choose the median of the three values as pivot.
	\item Pick $a_\ell$ with $\ell := a_{n-1} \mod n-1$ as pivot.
\end{itemize}
For each strategy, find an input such that quicksort takes $\Omega(n^2)$ time on that input.

\paragraph{Answer:}%
\newcommand*{\USED}[1]{{\color{blue}#1}}%
\newcommand{\HOLE}{{\rule[-.2ex]{1ex}{.4pt}}}%
\newcommand{\SWAP}{\leftrightarrow}%
\newcommand*{\argmedian}{\operatorname{argmedian}}%
\newcommand*{\set}[1]{\{#1\}}%
In the following, we describe how to obtain a counterexample, ``animate'' counterexample generation for a small size, and then ``animate'' quicksort on the resulting counterexample.
In the animations, we \USED{mark in blue} elements that are no longer ``in play''.
In quicksort animations, we simply sort any ``small'' subproblems that arise, so that, from one line to the next, we show only quicksort's ``big'' subproblems.
\begin{itemize}

	\item
	(midpoint):
	\textbf{Construction:}
	Create a sorted list of length $n$ and mark all positions $0,\ldots,n-1$ ``in play''.
	While there are positions in play, swap the minimal element that's still in play into the middle position (wrt the number of positions in play) and mark the middle position no longer in play:
	\[
		\begin{array}{cll}
		\multicolumn{3}{c}{\text{\emph{constructing a counterexample}}} \\
		\text{list} & \text{middle position} & \text{swap positions} \\
		1, 2, 3, 4, 5	&\floor{5/2} = 2 & 2 \SWAP 0\\
		3, 2, \USED{1}, 4, 5	&\floor{4/2} = 2 & 2 \SWAP 1 \\
		3, 4, \USED{1, 2}, 5	&\floor{3/2} = 1 & 1 \SWAP 0 \\
		4, \USED{3, 1, 2}, 5	&\floor{2/2} = 1 & 1 \SWAP 0 \\
		\USED{4, 3, 1, 2}, 5	&\floor{1/2} = 0 & 0 \SWAP 0 \\
		\USED{4, 3, 1, 2, 5} \\[\jot]
		\multicolumn{3}{c}{\text{\emph{running quicksort}}} \\
		\text{list} & \text{pivot position} \\
		4, 3, 1, 2, 5	& \floor{5/2} = 2 \\
		\USED{1}, 4, 3, 2, 5	&\floor{4/2} = 2 \\
		\USED{1, 2}, 4, 3, 5	&\floor{3/2} = 1 \\
		\USED{1, 2, 3}, 4, 5	&\floor{2/2} = 1 \\
		\USED{1, 2, 3, 4}, 5	&\floor{1/2} = 0 \\
		\USED{1, 2, 3, 4, 5}
		\end{array}
	\]
	Note that
	\begin{enumerate}
	\item Our algorithm terminates, as the number of positions in play decreases on each iteration.
	\item Since permutations are closed under swapping, it produces a permutation $\pi$ of the original sorted list.
	\item As $\text{quicksort}(\pi)$ runs, it always pivots on the minimal element.
	Thus its subproblems have size zero and $n-1$.
	The running time
	\begin{align*}
		T(0) &= \Theta(1) \\
		T(n) &= T(0) + n + T(n-1)
	\end{align*}
	is $\Theta(n^2$).

	\end{enumerate}

	\item
	(median of three):
	\textbf{Construction:}
	We use the same idea, but rather than ``poison'' (\ie take out of play by swapping in a minimal element) just the middle position, we posion both the first and middle positions:
	\[
		\begin{array}{cll}
		\multicolumn{3}{c}{\text{\emph{constructing a counterexample}}} \\
		\text{list} & \text{middle position} & \text{swap positions} \\
		1, 2, 3, 4, 5, 6, 7, 8	&\floor{8/2} = 4 & 0 \SWAP 0, 4 \SWAP 1 \\
		\USED{1}, 5, 3, 4, \USED{2}, 6, 7, 8 & \floor{6/2} = 3 & 0 \SWAP 1, 3 \SWAP 2 \\
		\USED{1, 3}, 5, 6, \USED{2, 4}, 7, 8 & \floor{4/2} = 2 & 0 \SWAP 0, 2 \SWAP 1 \\
		\USED{1, 3, 5}, 7, \USED{2, 4, 6}, 8 & \floor{2/2} = 1 & 0 \SWAP 0, 1 \SWAP 1 \\
		\USED{1, 3, 5, 7, 2, 4, 6, 8} \\[\jot]
		\multicolumn{3}{c}{\text{\emph{running quicksort}}} \\
		\text{list} & \text{pivot position} \\
		1, 3, 5, 7, 2, 4, 6, 8	& 4 \\
		\USED{1, 2}, 3, 5, 7, 4, 6, 8 & 3 \\
		\USED{1, 2, 3, 4}, 5, 7, 6, 8 & 2 \\
		\USED{1, 2, 3, 4, 5, 6}, 7, 8 & 0 \\
		\USED{1, 2, 3, 4, 5, 6, 7}, 8 & 0 \\
		\USED{1, 2, 3, 4, 5, 6, 7, 8}
		\end{array}
	\]
	At each stage, quicksort's pivot selection strategy examines the two smallest elements left in the array.
	Thus the subproblems created by quicksort have size one and $n - 2$.
	The running time
	\begin{align*}
		T(0) = T(1) &= \Theta(1) \\
		T(n) &= T(1) + n + T(n-2)
	\end{align*}
	is, again, $\Theta(n^2)$.

	\item (bizarre):
	\textbf{Construction:}
	Create an array of length $n$ and mark positions $0, \ldots, n-1$ ``in play''.
	While there are more than one positions in play,
	(i)~let $k$ be the number of positions,
	(ii)~drop $k$ into the in play position $p := k-1$,
	(iii)~drop $k-1$ into the in play position $p' := k \mod k-1 = 1$, and
	(iv)~mark positions $p$ and $p'$ out of play.
	For the final, in play position (if any) drop in $1$:
	\[
		\begin{array}{cl}
		\multicolumn{2}{c}{\text{\emph{constructing a counterexample}}} \\
		\HOLE, \HOLE, \HOLE, \HOLE, \HOLE, \HOLE, \HOLE \\
		\HOLE, \USED{6}, \HOLE, \HOLE, \HOLE, \HOLE, \USED{7} \\
		\HOLE, \USED{6, 4}, \HOLE, \HOLE, \USED{5, 7} \\
		\HOLE, \USED{6, 4, 2, 3, 5, 7} \\
		\USED{1, 6, 4, 2, 3, 5, 7} \\[\jot]
		\multicolumn{2}{c}{\text{\emph{running quicksort}}} \\
		\text{list} & \text{pivot position} \\
		1, 6, 4, 2, 3, 5, 7 & 1\\
		1, 4, 2, 3, 5, \USED{6, 7} & 1\\
		1, 2, 3, \USED{4, 5, 6, 7} & 1\\
		1, \USED{2, 3, 4, 5, 6, 7} & 1\\
		\end{array}
	\]
	As in the preceding case, the subproblems created by quicksort have size one and $n-2$, so the running time is $\Omega(n^2)$.
\end{itemize}

\subsection{Part b}

Assume for clarity that the input list contains no duplicates.
Quicksort achieves its worst-case running time of $\Theta(n^2) \subseteq \Omega(n^2)$ when the pivot is \emph{always} (the position of) the minimum element in its input.
One of the sublists on which quicksort recurses is empty, while the other is a list of length $n-1$.
Thus there are $\Theta(n)$ levels of recursion---each taking $\Theta(n)$ time on its own---for a total of $\Theta(n^2)$.

With this insight, our task is simple:
Permute a sorted list for a given pivot selection strategy, $f$, so that $f$ always selects the minimum.
\textbf{Algorithm to produce a worst-case input for a given pivot selection strategy:}
Create a sorted list of length $n$ and mark all items ``in play''.
While there are elements in play, swap the minimal element that's still in play into the pivot position, and mark the pivot as no longer in play.

Our algorithm terminates, as the number of items in play decreases on each iteration.
Since permutations are closed under swapping, it produces a permutation $\pi$ of the original sorted list.
By construction, this permutation forces $\text{quicksort}_{f}(\pi)$ to run in $\Theta(n^2)$ time, \emph{provided $f$ is deterministic}.

In the following, we ``animate'' our algorithm, draw a line, then (as a sanity check) animate quicksort.
(This would be easier with chalk!)
For both animations, we \USED{mark in blue} those elements in the list that are no longer ``in play''.
In the quicksort animations, of course, we only show the work done by one of the recursive calls.
For concreteness, we always start with the sorted list $1, 2, 3, 4, 5$.

	\[
		\begin{array}{cl}
		\text{list} & \text{pivot position} \\
		1, 2, 3, 4, 5	&\floor{5/2} = 2\\
		3, 2, \USED{1}, 4, 5	&\floor{4/2} = 2 \\
		3, 4, \USED{1, 2}, 5	&\floor{3/2} = 1 \\
		4, \USED{3, 1, 2}, 5	&\floor{2/2} = 1 \\
		\USED{4, 3, 1, 2}, 5	&\floor{1/2} = 0 \\
		\USED{4, 3, 1, 2, 5}	\\\hline
		4, 3, 1, 2, 5	& \floor{5/2} = 2 \\
		\USED{1}, 4, 3, 2, 5	&\floor{4/2} = 2 \\
		\USED{1, 2}, 4, 3, 5	&\floor{3/2} = 1 \\
		\USED{1, 2, 3}, 4, 5	&\floor{2/2} = 1 \\
		\USED{1, 2, 3, 4}, 5	&\floor{1/2} = 0 \\
		\USED{1, 2, 3, 4, 5}
		\end{array}
	\]

	\[
		\begin{array}{cl}
		\text{list} & \text{pivot position} \\
		1, 2, 3, 4, 5	&\argmedian\set{ 0\mapsto 1, 4\mapsto 5, \floor{5/2}=2\mapsto 3} = 2 \\
		3, 2, \USED{1}, 4, 5	&\argmedian\set{ 0\mapsto 3, 3\mapsto 5, \floor{4/2}=2\mapsto 4} = 2 \\
		3, 4, \USED{1, 2}, 5	&\argmedian\set{ 0\mapsto 3, 2\mapsto 5, \floor{3/2}=1\mapsto 4} = 1 \\
		4, \USED{3, 1, 2}, 5	&\argmedian\set{ 0\mapsto 4, 1\mapsto 5, \floor{2/2}=1\mapsto 5} = 1 \\
		5, \USED{3, 1, 2, 4}	&\argmedian\set{ 0\mapsto 5, 0\mapsto 5, \floor{1/2}=5\mapsto 5} = 0 \\
		\USED{5, 3, 1, 2, 4} \\\hline
		5, 3, 1, 2, 4	&\argmedian\set{ 0\mapsto 5, 4 \mapsto 4, \floor{5/2}=2\mapsto 1} = 4 \\
		3, 1, 2, \USED{4, 5}	&\argmedian\set{0 \mapsto 3, 2 \mapsto 2, \floor{3/2}=1 \mapsto 1} = 1 \\
		\USED{1}, 3, 2, \USED{4, 5} & \argmedian\set{0 \mapsto 3, 1 \mapsto 2, \floor{2/2}=1 \mapsto 2} = 1 \\
		\USED{1, 2}, 3, \USED{4, 5} & \argmedian\set{0 \mapsto 3, 0 \mapsto 3, \floor{1/2}=0 \mapsto 3} = 0 \\
		\USED{1, 2, 3, 4, 5}
		\end{array}
	\]

%%%%%%%%%%%%%%%%%%%%%%%%%%%% no conflicts, git! %%%%%

\section{Exercise 3 (10 pts)}

\noindent Consider a hash table of size $m$ to store $n \le  \frac{m}{2}$ elements. We use open addressing and assume a uniform family of hash functions.

\paragraph{a)} Show that for all $i = 1, \ldots, n$, the probability that the $i^{th}$ insertion requires strictly more than $k$ probes is at most $2^{-k}$.

\paragraph{Answer:}

By induction on the number of probes $k$. Consider the base case $k=0$.


\begin{align*}
\forall i \in \{1, \ldots, n\}, \\
& {prob}(X_i > 0) \le 2^{-k} = 2^0 = 1.
\end{align*}

By IH, assume that $\forall i \in \{1, \ldots, n\}, {prob}(X_i > k) \le 2^{-k}$. We know that:

\begin{align*}
& {prob}(X_i > k+1) && \text{By mutual exclusion of events} \\
& = {prob}(X_i > k) - {prob}(X_i = k+1) \\
& \le {prob}(X_i > k) && \text{[By IH]} \\
& \le 2^{-k}
\end{align*}

\paragraph{b)} Show that for $i = 1, \ldots, n$, the probability that the $i^{th}$ insertion requires more than $2 \log_2 n$ probes is at most $1/n^2$.

\paragraph{Answer:}

Assume $n$ is a power of $2$ (this makes no difference). We know that $\forall i \in \{1, \ldots, n\},$
${prob}(X_i > k) \le 2^{-k}$. Thus:

\begin{align*}
{prob}(X_i > 2 \log_2 n) & \le 2^{-2\log_2 n} \\
& \le \frac{1}{2^{2 \log_2 n}} \\
& \le \frac{1}{n^2}
\end{align*}


\noindent Let $X_i$ be a random variable counting the number of probes required in the $i^{th}$ insertion. You have shown above that ${prob}(X_i > 2 \log_2 n) \le 1/n^2$. Let $X = \max_{1 \le i \le n} X_i$ be the maximum number of probes required by any of the $n$ insertions.


\paragraph{c)} Show that $\{prob\}(X > 2 \log_2 n) \le 1/n$


\paragraph{Answer:}

\begin{align*}
{prob}(X > 2 \log_2 n) & \le {prob}(\bigvee\limits_{i=1}^n X_i > \frac{2 \log_2 n}{n}) \\
& = \sum\limits_{i=1}^n {prob}(X_i > \frac{2 \log_2 n}{n}) \\
& \le n \cdot \frac{1}{n^2} \\
& = \frac{1}{n}
\end{align*}

But I think we can obtain a tighter bound:

\begin{align*}
X & > 2 \log_2 n \\
\iff \max\limits_{1 \le i \le n} X_i & > 2 \log_2 n \\
\implies \sum\limits_{1 \le i \le n} X_i & > 2 \log_2 n \\
\iff \forall X_i, X_i & > \frac{2 \log_2 n}{n}
\end{align*}

That is, the event $\{X > 2 \log_2 n\} \subseteq \bigvee\limits_{i=1}^n \{X_i > \frac{2 \log_2 n}{n}\}$, which leads to a tighter bound on the probability.

\paragraph{d)} Conclude that the expected length of the longest probe sequence is $O(\log n)$.

\paragraph{Answer:}

\begin{align*}
E[X] & \le \int_1^n \! {prob}(X > t) \, \mathrm{d}t && \text{[Split into short and long sums]} \\
& = \int_1^{2 \log_2 n} \! {prob}(X > t) \, \mathrm{d}t + \int_{2 \log_2 n}^n \! {prob}(X > t) \, \mathrm{d}t && \text{[Approximate short sum with probability = 1]} \\
& \le \int_1^{2 \log_2 n} \! \, \mathrm{d}t + \int_{2 \log_2 n}^n \! {prob}(X > t) \, \mathrm{d}t && \text{[Approximate long sum, since ${prob}(X > t)$ decreases]} \\
& \le \int_1^{2 \log_2 n} \! \, \mathrm{d}t + \int_{2 \log_2 n}^n \! {prob}(X > 2 \log_2 n) \, \mathrm{d}t && \text{[By result of question (c)]}\\
& \le \int_1^{2 \log_2 n} \! \, \mathrm{d}t + \int_{2 \log_2 n}^n \! \frac{1}{n} \, \mathrm{d}t \\
& \le (2\log_2 n - 1) + \frac{1}{n} (n - 2\log_2 n) \\
& \le 2\log_2 n - \frac{2\log_2 n}{n} \\
& \le 2\log_2 n \\
& = O(\log n)
\end{align*}


\section{Exercise 4 (10 pts)}

\noindent Consider the following problem: for some group of people, we know a list of their $n$ favorite songs, and we want to identify people with similar taste quickly. For each person, we create a Bloom filter, using the same table size $m$ and the same set of $k$ hash functions (from a 1-universal family). For any two people. determine the expected number of bits where the Bloom filters differ as a function of $n, m, k$, and $s$, the number of common songs in their lists. Explain how to estimate $s$ after measuring the numbers of differing bits of two people. What is the advantage of this approach in practice compared to direct comparison of the lists of songs?

\paragraph{Answer:}

\end{document}
