\section{Stable Matching}

Discussed by Gale, Shapley in 1962 (context of job markets): This problem is also related the allocation of workers in hospitals.

Each company a strict preference over applicants. Applicant has a strict preference over companies.

Is there a solution that is self-enforcing?

\begin{mydefinition}[Self-enforcing]
A matching is self-enforcing if for every applicant A and company C, such that A is not assigned to C, either:
\begin{itemize}
\item C prefers every one of its accepted applicants over A, or 
\item A prefers his current situation over working for C.
\end{itemize}
\end{mydefinition}



\section{Formal Model}

\begin{itemize}
\item Set X of man, 	$n = |X|$

\item Set Y of woman,	$m = |Y|$

\item Each $x \in X$ has a strict preference order $>_x$ over $Y$.

\item Each $y \in Y$ has a strict preference order $>_y$ over $X$.

\item Matching $M \subseteq X \times Y$ is a set of ordered pairs such that
every agent appears in at most one pair.
\end{itemize}

A blocking pair $(x,y)$ is a pair such that either $x$ or $y$ prefers some other person than its current match. A stable matching does not have blocking pairs.

\textbf{Problem:} Does there always exist a stable matching? Can we compute it in polynomial time?

Note: There may be solutions that are unfair, that is, they satisfy only one side (companies or workers).

\section{Deferred Acceptance Algorithm}

\textbf{Main idea}: Starts with an empty set of edges.

\begin{algorithmic}[1]
    \State{Initially, all $x \in X$ and all $y \in Y$ are free.}
    \While{$\exists$ free man $x$ that has not proposed to all women}
        \State{Let $y$ be the highest-ranked proposed women for $x$ that he hasn't proposed to
}
        \State{$x$ proposes to $y$}
        \If{$y$ is free}
            \State{$(x,y)$ gets engaged}
        \Else
            \State{$y$ currently engaged to $x'$}
            \If{$x >_y x'$}
                \State{$(x,y)$ gets engaged and $x'$ becomes free}
            \EndIf
        \EndIf
    \EndWhile
\end{algorithmic}

\begin{theorem}
The DA algorithm terminates after $n \cdot m$ iterations.
\end{theorem}
\begin{proof}
Some straightforward properties:
\begin{enumerate}
\item If $y$ becomes engaged, she will be matched in the end. The engagement partners are improving with respect to her preference $>_y$ over time.
\item $x$ proposes in decreasing order of $>_x$.
Thus, number of iterations is upper bounded by $n \cdot m$.
\end{enumerate}

Let $y >_x y'$ and $x >_y x'$. By contradiction, suppose in the final matching there is a BP, that is, we have a matching $M=\{(x,y'), (x', y)\}$.

- Note that $y >_x y'$, so by algorithm, $x$ proposed to $y$.

- Then $x$ proposed to $y'$, but $x$ becomes free afterwards.

- $y$ had a better engaged partner $x' >_y x$.

Contradiction.
\end{proof}

\subsection{Formalisms}

\begin{enumerate}

\item Woman $y$ is a \emph{valid partner} for a man $x$ if there exists a SM $M$ with $(x,y) \in M$.

\item Valid partner $y$ is the \emph{best valid partner} for man $x$ if for every valid partner $y' \neq y$, we have that $y >_x y'$.

\item Similar for \emph{worst valid partner}.
\end{enumerate}

\textbf{Notation:} $y = best(x)$ and $best(x) = x$ if $x$ has no valid partner.

We define the optimal matching as follows:

$$M^+ = \{ (x, best(x) | x \in X \wedge best(x) \neq x\}$$

\begin{theorem}
Every execution of the DA algorithm results in $M^+$.
\end{theorem}
\begin{proof}
Contradiction. Assume there is an execution $E$ of the DA algorithm resulting in a SM $M$, where some man is not matched to his best valid partner. This means that this man was rejected by a better valid partner during $E$.

Consider the first time that this happens, i.e., a man $x$ is rejected by a valid partner.

We have $(x',y)$, with $x' >_y x$.

\begin{enumerate}
\item $x$ proposed in decreasing order of preference, $y$ is ${best}(x)$.
\item $y$ rejects $x$, so $y$ is engaged to $x'$, with $x' >_y x$.
\item There exists Sm M with $(x,y) \in M$. Let $(x', y') \in M$.
\item Since rejection by $y$ was first in $E$, then $x'$ is not rejected by any valid partner when engaged to $y$. Since $y'$ is a valid partner of $x'$, $y >_x y'$. Then $(x', y)$ is a BP for $M$. But we assumed that $M$ is stable! Contradiction.
\end{enumerate}
\end{proof}

\begin{theorem}
In $M^+$, every woman is matched to her worst valid partner (if any).
\end{theorem}
\begin{proof}
Suppose $(x,y) \in M^+$ such that $x$ is not the worst valid partner of $y$.
Then there exists an SM $M'$ with $(x', y) \in M'$ and $x >_y x'$.
In $M'$ we have $(x, y') \in M'$. $y = {best}(x)$ and $y'$ is a valid partner for $x$. So $y >_x y'$. Then $(x,y)$ is a BP in $M'$. Contradiction.
Similar if $x$ is unmatched in $M'$.
\end{proof}